\texttt{Taxonomic supertree construction with}\texttt{\emph{ Incertae sedis}}\texttt{ taxa}

### 2018-06-18 doc/incertae_sedis.tex
Updating section 2.6

After we have attached taxon names to the synthesis tree, we would
like to interpret the position of these names in terms of placing
\emph{incertae sedis }taxa in a revised taxonomy.
This would allow us
to interpret the synthesis tree as saying that phylogenetic
information has (for example) placed genus A within family B, or
perhaps outside of all named families.
The simplest approach to
placement involves noting whenever a taxon $B$ is a descendant of a
taxon $A$ on the named synthesis tree but not the taxonomy tree.
For
example, in Figure \ref{fig:Handling-incertae-sedis}, taxon $C$ is a
descendant of taxon $A$ in the synthetic tree 
(Fig.\ref{fig:Handling-incertae-sedis}d), but not the taxonomy 
(Fig.\ref{fig:Handling-incertae-sedis}b).
Thus, we could say that the synthetic tree places
$C$ within $A$. 
%Furthermore, the synthetic tree additionally places
% $C$ within $AB$.

In general, after naming the clades in the supertree, the
    taxonomic placement of an \emph{incertae sedis} taxon
    can be detected if the most tip-ward named ancestor of
    the clade in the supertree was not an ancestor in the
    taxonomic tree.
Note that if an \emph{incertae sedis} taxon is broken
    during the supertree algorithm, it can be ambiguous whether
    or not all of the members should be placed in a new
    parent taxon.
See, for example, Figure \ref{fig:Broken-incertae-sedis}.
Here taxon $T1$ is a descendant of $R$ in the
synthesis tree, but not the taxonomy.
Thus $T1$ is placed within $R$; taxon
 $T2$ is similarly placed with in $R$.
$T3$ is not mentioned in either input tree, and the 
    only tree that speaks to its placement is the taxonomic input.
However, the taxon $T$ cannot be monophyletic based on the
     two input trees.
Thus the correct placement of $T3$ is uncertain.

The ambiguity of where to place $T3$ can also be compared to the
    behavior of maximum compatibility tree estimation and parsimony.
The matrix representation context the taxonomic tree's
    character representation of taxon $T$ would include $T1$, $T2$ and $T3$
    all being assigned state 1 and all of the other leaves being assigned 
    state 0.
This character is not compatible with the solution tree. 
The compatability approach would be to conclude that the character encoding 
    that taxon was incorrect and should be ignored.
Since, no other character speaks to the position of $T3$, it would fall to being
    a child of the root of the tree.
In a similar result, the BUILD algorithm would place $T3$ as a child of the root.

In a parsimony view, while the character cannot be mapped onto the solution in
    a homoplasy-free manner, one might hope that the character data that the
    taxonomy was based on was not completely unreliable.
Placing of $T3$ as either sister to $T1$ or $T2$ could explain the character
    representation of taxon $T$ with only two acquisitions of the state 1.
Other placements of $T3$ would require at least three 0 to 1 changes (or a reversion
    to state 0).
Placing $T3$ at the MRCA 
    of the other members of $T$ corresponds
    to handling of ``rogue'' taxa in an Adam's consensus tree \citep{Adams1972} method of the two
    most parsimonious placements of $T3$ under an irreversible parsimony model.

>>>>

\subsubsection{Revision of unbroken \emph{incertae sedis} taxa}
After we have attached taxon names to the synthesis tree, we would
like to interpret the position of these names in terms of placing
\emph{incertae sedis }taxa in a revised taxonomy.
This would allow us
to interpret the synthesis tree as saying that phylogenetic
information has (for example) placed genus A within family B, or
perhaps outside of all named families.
The simplest approach to
placement involves noting whenever a taxon $B$ is a descendant of a
taxon $A$ on the named synthesis tree but not the taxonomy tree.
For
example, in Figure \ref{fig:Handling-incertae-sedis}, taxon $C$ is a
descendant of taxon $A$ in the synthetic tree 
(Fig.\ref{fig:Handling-incertae-sedis}d), but not the taxonomy 
(Fig.\ref{fig:Handling-incertae-sedis}b).
Thus, we could say that the synthetic tree places
$C$ within $A$.

In general, after naming the clades in the supertree, the
    taxonomic placement of an \emph{incertae sedis} taxon
    can be detected if the name of the \emph{incertae sedis} taxon
    survives in the solution tree, and the most tip-ward named
    ancestor of the clade in the supertree was not an ancestor in the
    taxonomic tree.  In complex scenarios such as Figure
    \ref{fig:Placing-one-incertae}c, it may be necessary to
    walk the solution tree in pre-order and revise the taxonomy
    incrementally, in order to (for example) place V within WX
    before placing W within V.

\subsubsection{Revision of broken \emph{incertae sedis} taxa}
When an \emph{incertae sedis} taxon is broken and its name does not
occur on the solution tree, taxonomic revision becomes considerably
more complex.  For example, it can be ambiguous whether or not all of
the members of a broken taxon should be placed in a new parent
taxon. See, for example, Figure \ref{fig:Broken-incertae-sedis}. Here
taxon $T1$ is a descendant of $R$ in the synthesis tree, but not the
taxonomy. Thus $T1$ is placed within $R$; taxon  $T2$ is similarly
placed with in $R$. $T3$ is not mentioned in either input tree, and
the  only tree that speaks to its placement is the taxonomic
input. However, the taxon $T$ cannot be monophyletic based on the two
input trees.  Thus the correct placement of $T3$ is unclear.
Furthermore, the taxon name for $T$ does not occur on the solution tree.

The ambiguity of where to place $T3$ can explored by comparing the
behavior of maximum-compatibility tree estimation and parsimony-based
scoring schemes such as MRP. The taxonomic tree's character
representation of taxon $T$ would include $T1$, $T2$ and $T3$ 
all being assigned state 1 and all of the other leaves being assigned 
state 0.  This character is not compatible with the solution tree. 
The compatibility approach would be to conclude that the character
(or, equivalently, the corresponding rooted split) encoding that taxon was incorrect and should be completely ignored.
Since, no other character speaks to the position of $T3$, it would
fall to being  a child of the root of the tree.  This seems to
contradict the intuition that $T3$ should be placed at the MRCA of
$T1$ and $T2$.
\bdrcomment{This is not really a feature of BUILD, but of our choice to
completely throw all all triplets from incompatible splits.}

Using a parsimony-based scoring scheme, the information from the
character that encodes taxon $T$ is not completely ignored, even though it
is contradicted by an input tree.  Placing $T3$ as either sister to
$T1$ or $T2$ could explain the character representation of taxon $T$
with only two acquisitions of the state 1. Other placements of $T3$
would require at least three 0 to 1 changes (or a reversion to state
0). Placing $T3$ at the MRCA of the other members of $T$ corresponds
to handling of ``rogue'' taxa in an Adam's consensus tree
\citep{Adams1972} method of the two  most parsimonious placements of
$T3$ under an irreversible parsimony model.

Our current supertree algorithm places $T3$ at the MRCA of $T1$ and
$T2$ if $T3$ is a taxonomy-only taxon, but places $T3$ at the root
of the tree if it occurs in an input tree.  This inconsistent behavior
is not restricted to cases that contain \emph{incertae sedis} taxa, but is a
general problem with compatibility-based algorithms that do not retain
partial information from incompatible splits.  Such information could
include compatible triplets that comprise parts of the incompatible split.
The inconsistency results from the fact that our sub-problem solver
implements compatibility-based consensus and completely discards
incompatible splits, while the unpruner implements an \emph{ad hoc}
rule to place pruned taxonomy-only taxa at the MRCA of taxa that occur
in input trees.  It would be desirable to update the sub-problem
solver to be consistent with the unpruner, so that we could treat the
pruning and unpruning steps as an optimization that has no effect on
semantics.

It is possible to update our compatability-based supertree algorithm
to achieve the intuitive result of placing $T3$ at the MRCA of $T1$
and $T2$ without resorting to a parsimony-based scoring scheme. We
note that the split $T1\,T2\,T3\,|\bullet\,U\,S\,R1\,R2$ implies
many rooted triplets, and that we need not discard all of these
triplets because some of them are contradicted by input trees.  A
simple method of retaining partial information from incompatible
rooted splits would be to note that the MRCA of $T1$, $T2$ and must
also have $R1$ and $R2$ as descendants.  We would then remove these
additional taxa from the exclude group, obtaining the partial split
$T1\,T2\,T3\,|\bullet\,U\,S$.  Adding this split to the solution
tree places $T3$ at the MRCA of $T1$ and $T2$ as desired.  While this
approach places all members of $T$ inside $R$, it does not
place a monophyletic $T$ within $R$, since $T$ is not monophyletic.
Placing the name for $T$ at the MRCA of its members would make $T$ a
synonym for $R$.
